Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-09-20 manuscript/manuscript.tex

PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a software for the analysis of molecular dynamics (MD) simulations using Markov state models~\cite{schuette-msm,singhal-msm-naming} (MSMs).
The package is written in Python (\url{http://python.org}), relies heavily on NumPy/SciPy~\cite{numpy,scipy}, and is compatible with the scikit-learn~\cite{sklearn} framework for machine learning.

@@

A commonly used method for dimensionality reduction, TICA, is a particular implementation of the VAC.
To apply TICA, we need to compute instantaneous ($\mathbf{C}(0)$) and time-lagged ($\mathbf{C}(\tau)$) covariance matrices with elements
\begin{eqnarray}
c_{ij}(0) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t) \right\rangle_t \\
c_{ij}(\tau) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t + \tau) \right\rangle_t,
\end{eqnarray}
where $\tilde{x}_i(t)$ denotes the $i^\textrm{th}$ feature at time $t$ after the mean has been removed.
Then, we can solve the generalized eigenvalue problem
\begin{equation}
\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i \,, \quad i=1,\dots,n,
\end{equation}
to obtain independent component directions $\mathbf{u}_i$ which approximate the reaction coordinates of the system,
where the pairs of eigenvalues and independent components are sorted in descending order,
and we define a cumulative kinetic variance fraction
\begin{equation}
c_d = \frac{\sum_{i=2}^d \lambda_i^2(\tau)}{\textrm{TKV}},
\end{equation}
where
\begin{equation}
\textrm{TKV} = \sum\limits_{i=2}^n \lambda_i^2(\tau)
\end{equation}
is the total kinetic variance explained by all $n$ features.

If we further scale the independent components $\mathbf{u}_i$ by the corresponding eigenvectors $\lambda_i(\tau)$,
we obtain a \emph{kinetic map} which is the default behavior in PyEMMA.

Note, though, that TICA requires the data to be in equilibrium.
To use TICA with nonequilibrium data, we can either symmetrize the time-lagged covariance matrix $\mathbf{C}(\tau)$
or apply a Koopman reweighting~\cite{vamp-preprint}.
For short trajectories and nonequilibrium data we generally recommend to use VAMP~\cite{vamp-preprint}.

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{figure_1}
\caption{The HMM transition matrix $\tilde{\mathbf{P}}(\tau)$ propagates the hidden state trajectory $\tilde{s}(t)$ (orange circles) and, at each time step $t$, the emission into the observable state $s(t)$ is governed by the emission probabilities $\bm{\chi}\left( s(t) \middle| \tilde{s}(t) \right)$.}
\label{fig:hmm-scheme}
\end{figure}

@@

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space.
Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
and gives rise to our observed data, e.g., the discretized trajectories (see Fig.~\ref{fig:hmm-scheme}).
This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
our molecular dynamics trajectories.

To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
in practice, a timescale separation of $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
\begin{equation}
\mathbf{P}(k\tau) \approx \bm{\Pi}^{-1} \bm{\chi}^\top \tilde{\bm{\Pi}} \tilde{\mathbf{P}}^k(\tau) \, \bm{\chi}.
\end{equation}
Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m<n$ hidden states' stationary probabilities,
$\tilde{\mathbf{P}}(\tau)$ is a transition matrix between the $m<n$ hidden states at lag time $\tau$,
and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
where each row encodes the emission probabilities into the $n$ microstates conditioned on being in the corresponding hidden state~\cite{noe-proj-hid-msm}.

>>>>

PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a software for the analysis of molecular dynamics (MD) simulations using Markov state models~\cite{schuette-msm,singhal-msm-naming,noe2007jcp,chodera2007jcp,buchete-msm-2008} (MSMs).
The package is written in Python (\url{http://python.org}), relies heavily on NumPy/SciPy~\cite{numpy,scipy}, and is compatible with the scikit-learn~\cite{sklearn} framework for machine learning.

@@

A commonly used method for dimensionality reduction, TICA, is a particular implementation of the VAC.
To apply TICA, we need to compute instantaneous ($\mathbf{C}(0)$) and time-lagged ($\mathbf{C}(\tau)$) covariance matrices with elements
\begin{eqnarray}
c_{ij}(0) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t) \right\rangle_t \label{eq:c0}\\
c_{ij}(\tau) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t + \tau) \right\rangle_t, \label{eq:ct}
\end{eqnarray}
where $\tilde{x}_i(t)$ denotes the $i^\textrm{th}$ feature at time $t$ after the mean has been removed.
By default, PyEMMA estimates (\ref{eq:c0},\ref{eq:ct}) using symmetrization~\cite{tica}.
This symmetrization induces a significant bias when using non-equilibrium data from short trajectories~\cite{hao-variational-koopman-models}.
As an alternative, the so-called Koopman reweighting estimator is available which avoids this bias,
but comes at the cost of a large variance~\cite{hao-variational-koopman-models}.

After estimating the covariance matrices, TICA solves the generalized eigenvalue problem
\begin{equation}
\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i \,, \quad i=1,\dots,n,
\end{equation}
to obtain independent component directions $\mathbf{u}_i$ which approximate the reaction coordinates of the system,
where the pairs of eigenvalues and independent components are sorted in descending order.
A way to measure the contribution of each independent component to the kinetics
is obtained by the kinetic distance~\cite{kinetic-maps}
which assigns a cumulative variance fraction to the first $d$ independent components:
\begin{equation}
c_d = \frac{\sum_{i=2}^d \lambda_i^2(\tau)}{\textrm{TKV}},
\end{equation}
where
\begin{equation}
\textrm{TKV} = \sum\limits_{i=2}^n \lambda_i^2(\tau)
\end{equation}
is the total kinetic variance explained by all $n$ features.

If we further scale the independent components $\mathbf{u}_i$ by the corresponding eigenvectors $\lambda_i(\tau)$,
we obtain a \emph{kinetic map}~\cite{kinetic-maps} which is the default behavior in PyEMMA.

Note, though, that TICA requires the dynamics to be simulated at equilibrium conditions.
To use TICA with nonequilibrium MD, e.g., subject to external forces,
or simply to perform dimension reduction on short trajectory data without worrying about reweighting,
we recommend to use VAMP~\cite{vamp-preprint}.

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{figure_1}
\caption{The HMM transition matrix $\tilde{\mathbf{P}}(\tau)$ propagates the hidden state trajectory $\tilde{s}(t)$ (orange circles) and, at each time step $t$, the emission into the observable state $s(t)$ (cyan circles) is governed by the emission probabilities $\bm{\chi}\left( s(t) \middle| \tilde{s}(t) \right)$.}
\label{fig:hmm-scheme}
\end{figure}

@@

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
and gives rise to our observed data, e.g., the ($n$~states) discretized trajectories $s(t)$.
This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
our molecular dynamics trajectories.

To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ timescale;
in practice, a timescale separation of $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
The HMM then consists of a transition matrix $\tilde{\mathbf{P}}(\tau)$ between $m<n$ hidden states
and a row-stochastic matrix ($\bm{\chi}$) of probabilities $\chi\left( s \middle| \tilde{s} \right)$
to emit the discrete state $s$ conditional on being in the hidden state $\tilde{s}$.

We can further compute a reversal of the emission matrix $\bm{\chi}\in\mathbb{R}^{m \times n}$:
the membership matrix $\mathbf{M}\in\mathbb{R}^{n \times m}$ which encodes
a fuzzy assignment of each of the $n$ observable microstates $s$ to the $m$ hidden states $\tilde{s}$ and,
thus, defines the \emph{coarse graining} of microstate.
